%!TEX root = thesis.tex
\newpage
\paragraph{Abstract} 
Transactions in non-relational databases, are a rare feature, but a useful and desirable one if present, such as Light Weight Transactions (\lwt) in the Cassandra.
The problem with \lwt is lack of a multi partition operations, which limits the practical applications, and puts the burden of maintaining the data consistency in other ways on the users.
The proposed multi-partition transactions algorithm removes the burden, supports transactions spanning many rows with read-committed isolation, and has a scalable design without a single point of failure.
We have achieved this by using \paxos algorithm for the distributed consensus, private memtables for isolation, and other techniques, which reduce memory footprint, and increase scalability.

% Multiple-inheritance hierarchies are data structures of a high importance in the area of
% the artificial intelligence. 
% The problem with them is lack of comprehensive and efficient methods of indexing, which
% limits the practical applications.
% The proposed hybrid indexing method can adapt to the particular topology of the data
% structure being processed. 
% We have achieved this by tuning the proportion between the member encoding methods to gain
% compactness of the index and fast responses of the search requests.
% Moreover, we have found that combining a few methods
% together preserves and even improves the ability to encode the changes incrementally.

% We have devised two member encoding methods: the first that bases on inheritance of some features and the second that
% bases on numbering schemes. The correctness of the developed solutions has been proved and their performance has been
% evaluated in the testing environment. The results show the characteristics of the hybrid index and prove its
% efficiency.

\paragraph{Keywords:} indexing transitive relations, multiple-inheritance hierarchy, reachability testing

\begin{center}
\vspace*{\baselineskip}
    \fontseries{b}\fontsize{15pt}{18pt}\selectfont
      Transakcje dla wielu partycji w Cassandrze  \\
\end{center}
\paragraph{Streszczenie} Hierarchie dziedziczenia wielokrotnego są strukturami danych o szczególnym znaczeniu w
obszarze zastosowań sztucznej inteligencji. Problemem, który ogranicza ich wykorzystanie w praktyce jest brak
uniwersalnej i efektywnej metody ich indeksowania. Zaproponowana hybrydowa metoda indeksowania dostosowuje się do
topologii przetwarzanej struktury danych. Osiągnięto to po przez dostrajanie proporcji pomiędzy składowymi metodami
kodowania, tak aby indeks był kompaktowy i oferował szybki dostęp do danych. Ponadto, połączenie komplementarnych metod
kodowania nie powoduje utraty zdolności do kodowania inkrementalnego, a nawet ją wspomaga.

% W ramach pracy stworzone zostały dwie składowe metody kodowania: jedna bazująca na dziedziczeniu cech, druga bazująca na
% schematach numerowania. Poprawność opracowanych rozwiązań została udowodniona, a wydajność sprawdzona w środowisku
% testowym. Wyniki obrazują charakterystykę hybrydowej metody indeksowania i dowodzą jej efektywności.

\paragraph{Słowa kluczowe:} indeksowanie relacji przechodnich, hierarchia dziedziczenia wielokrotnego, sprawdzanie
istnienia ścieżki w grafie

\pagenumbering{gobble}
