%!TEX root = ../thesis.tex

\subsection{Conflicting transactions}\label{sec:impl:conflictFunctions}
In order for node \node{i} to know whether two transactions are conflicting it requires a comparison of their mutations and a check whether they mutate the same keys. However keys are unknown to transaction items \txItems, thus
given transaction states \txStateOne and \txStateTwo such check requires sending network messages to replicas, which compare mutations in private transaction storage of \txStateOne and \txStateTwo. 
Distributed algorithms should minimize round trips, because they cost resources and decrease performance \cite{rotem2006fallacies}.

Local conflict functions, which use only transaction state \txState as it is (without accessing mutations \mutations on all replica nodes affected by transaction \nodesTx) do not require network messaging, but as a tradeoff may raise false positives, as presented in Figure \ref{fig:conflictFunctions}. 

\begin{figure}
\centering
\begin{tabular}{|p{3cm}|p{11cm}|}
        \toprule
        Function name & Description \\ \midrule
        Conflict on the same transaction item \txItem &
If transactions \transactioni{1}, \transactioni{2} modify a row with key $k_{1}$ in table $t_{1}$ then both transactions have the same transaction item \txItem. Given transaction states of two transactions, transaction items can be checked for intersection.
	If they intersect then transactions are considered conflicting. 
	False positives occur when transactions modify a row with key $k_{1}$, but \transactioni{1} modifies column $c_{1}$ and \transactioni{2} modifies column $c_{2}$. 
	Transaction items for the same partition $k_p$ but different $k_c$ are not distinguished, thus function can yield conflict when there is none. \\
	Conflict on the same table & Function marks transactions as conflicting if they have \txItem for the same table. It greatly increases number of conflicts, but reduces number of Paxos rounds since all conflicting transactions participate in the same Paxos round.
		False positives occur when transactions modify different keys $k_{1}, k_{2}$ in the same table.  \\
	
	Conflict on the same token range slice & 
         Function slices token range into $s$ slices, where $s$ is a configurable parameter. Function takes \txItem and finds slice number for token in \txItem that falls into sliced token range. If two transactions have transaction items which fall into the same slice, then transactions are considered conflicting.
		Fuction allows to have configurable number of concurrent Paxos rounds since there will be at most, as many rounds, as the number of slices. False positives are expected result of the function.
         \\ \bottomrule
      \end{tabular}     
     \caption{Conflict functions}
     \label{fig:conflictFunctions}
\end{figure}


Conflict functions presented in Figure \ref{fig:conflictFunctions} eliminate network messages and put bounds on number of concurrent transactions, but with a cost of false positives. In principle the transaction index \txIndex can use any conflict function, which satifies its properties, thus it is possible to develop a new function, which may use network messages to detect conflicts, but such function would decrease performance of the solution. 



\subsection{Configurable conflict functions}
The implementation provides configuration of conflict functions per column family, thus supporting different choice of conflict functions for different tables depending on intended workload.

Per table configuration is an additional feature to the algorithm implemented using column family metadata, which can be changed dynamically at runtime of the cluster by executing \code{ALTER TABLE} commands. Listing \ref{lst:configureConflictFunctions} shows how the conflict function is set to resolve conflict on the same token range slice number, which puts bounds on number of concurrent transactions in table \code{sliced_counters} to $2000$ transactions. 

\begin{lstlisting}[style=outcode,label={lst:configureConflictFunctions},caption={Configuring conflict function on Cassandra table}]
CREATE TABLE counters.sliced_counters(
  id uuid, counter1 int, counter2 int, 
  PRIMARY KEY (id)
);
ALTER TABLE counters.sliced_counters 
WITH extensions = { 
   'transaction_conflict_bounds' : 'TOKEN_RANGE_SLICES', 
   'token_range_slices' : '2000' 
 };
\end{lstlisting}

Conflict functions are available with the following names:
\begin{enumerate*}[label=\alph*)]
\item \code{COMMON_TX_ITEMS} -- conflict on the same transaction items, used as the default conflict function,
\item \code{TOKEN_RANGE_SLICES} -- conflict on the same token range slice number. Function has additional configurable parameter \code{int token_range_slices}
\item \code{ONE_FOR_ALL} -- conflict on the same table.
\end{enumerate*}
